---
title: "flowPloidy: Overview"
author: "Tyler Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flowPloidy: Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: flowPloidy.bibtex
---

# Installation

`flowPloidy` depends on the Bioconductor package `flowCore`. You will have to install this dependency before we get started (this requires an internet connection):

```{r, eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite()
biocLite("flowCore")
```

You only need to do this once, not each time you use `flowPloidy`. Once you've done so, you can install `flowPloidy`, and the remaining dependencies will be automatically installed for you. `flowPloidy` is currently available from [BitBucket](https://bitbucket.org/tws/flowPloidy), and you can install it from there using the `devtools` package (which must already be installed):

```{r, eval=FALSE}
library(devtools)
install_bitbucket("tws/flowPloidy")
```

# Getting Started
Once all the necessary libraries are installed, we're ready to start working with our FCS data.

This is a test citation [@bagwell_1993].

```{r}
x <- 1:10
y <- 2:11
plot(x, y)
```

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

# References
